\textbf{SPECFEM3D}\\ \textbf{User Manual}

### 2011-01-31 doc/USER_MANUAL/manual_SPECFEM3D.tex
updates check_mesh_quality_CUBIT_Abaqus/ and manual; updates paper_draft/

The SPECFEM3D software package comes in a gzipped tar ball. In the
directory in which you want to install the package, type

\begin{lyxcode}
{\small tar~-zxvf~SPECFEM3D\_V2.0.0.tar.gz}{\small \par}
\end{lyxcode}
The directory \texttt{\small SPECFEM3D} will then contain
the source code. To configure the software for your system, run the
\texttt{configure} shell script. This script will attempt to guess
the appropriate configuration values for your system. However, at
a minimum, it is recommended that you explicitly specify the appropriate
command names for your Fortran90 compiler and MPI package:

@@

\subsubsection*{Checking the mesh quality}
The quality of the mesh may be inspected more precisely based upon the serial code
in the file  ~\\
\texttt{check\_mesh\_quality\_CUBIT\_Abaqus.f90}.
Running this code is optional because no information needed by the
solver is generated.

Prior to running and compiling this code, you have to export your mesh in CUBIT to an ABAQUS (.inp) format.
You also have to determine a number of parameters of your mesh, such as the number of nodes and number of elements
and modify the header of the \texttt{check\_mesh\_quality\_CUBIT\_Abaqus.f90} source file. Then,
in the directory \texttt{src/check\_mesh\_quality\_CUBIT\_Abaqus}, type
\begin{lyxcode}
{\small ./compile\_all.csh}{\small \par}
\end{lyxcode}
and use
\begin{lyxcode}
{\small xcheck\_mesh\_quality\_CUBIT\_Abaqus}{\small \par}
\end{lyxcode}
to generate an AVS output file (\texttt{\small AVS\_meshquality.inp}
in AVS UCD format) or OpenDX output file (\texttt{\small DX\_meshquality.}~\\
\texttt{\small dx}) that can be used to investigate mesh quality,
e.g, skewness of elements and a Gnuplot histogram (\texttt{\small mesh\_quality\_}~\\
\texttt{\small histogram.txt}) that can be plotted with gnuplot (type
`\texttt{\small gnuplot plot\_mesh\_quality\_histogram.gnu}'). The
histogram is also printed to the screen.
Analyze that skewness histogram of mesh elements to make sure no element has a skewness above approximately 0.75, otherwise the mesh is of poor quality (and if even a single element has a skewness value above 0.80, then you must definitely improve the mesh).
If you want to start designing
your own meshes, this tool is useful for viewing your creations. You
are striving for meshes with elements with `cube-like' dimensions,
e.g., the mesh should contain no very elongated or skewed elements.

@@

When you are ready to compile,
in the main directory type '\texttt{make decompose\_mesh\_SCOTCH}'.
If all paths and flags have been set correctly, the executable \texttt{bin/xdecompose\_mesh\_SCOTCH} should be produced.

@@

In the main directory type `\texttt{make meshfem3D}'. If all paths and flags
have been set correctly, the mesher should now compile and produce
the executable \texttt{bin/xmeshfem3D}. Please note that
\texttt{xmeshfem3D} must be called directly from the \texttt{bin/} directory,
as most of the binaries of the package.

@@

After using \texttt{xmeshfem3D} or \texttt{xdecompose\_mesh\_SCOTCH}, the next step in the workflow is to compile \texttt{xgenerate\_databases}. This program is going to create all the missing information needed by the SEM solver.

@@

In the main directory type '\texttt{make generate\_databases}'. Input for the program is provided through
the main parameter file \texttt{Par\_file}, which resides in the subdirectory \texttt{in\_data\_files/}. Please note that
\texttt{xgenerate\_databases} must be called directly from the \texttt{bin/} directory, as most of the binaries of the package.

@@

Now that you have successfully generated the databases, you are ready to compile
the solver. In the main directory type `\texttt{make specfem3D}'.
Please note that \texttt{xspecfem3D} must be called directly from
the \texttt{bin/} directory, as most of the binaries of the package.

@@

\begin{itemize}
\item Set the \texttt{time shift} parameter equal to $0.0$ (the solver
will not run otherwise.) The time shift parameter would simply apply
an overall time shift to the synthetics, something that can be done
in the post-processing (see Section \ref{sec:Process-data-and-syn}).
\item For point-source simulations (see finite sources, page \pageref{To-simulate-a})
we recommend setting the source half-duration parameter \texttt{half
duration} equal to zero, which corresponds to simulating a step source-time
function, i.e., a moment-rate function that is a delta function. If
\texttt{half duration} is not set to zero, the code will use a Gaussian
(i.e., a signal with a shape similar to a `smoothed triangle', as
explained in \citet{KoTr02a} and shown in Fig~\ref{fig:gauss.vs.triangle})
source-time function with half-width \texttt{half duration}. We prefer
to run the solver with \texttt{half duration} set to zero and convolve
the resulting synthetic seismograms in post-processing after the run,
because this way it is easy to use a variety of source-time functions
(see Section \ref{sec:Process-data-and-syn}). \citet{KoTr02a} determined
that the noise generated in the simulation by using a step source
time function may be safely filtered out afterward based upon a convolution
with the desired source time function and/or low-pass filtering. Use
the serial code \texttt{convolve\_source\_timefunction.f90} and the
script \texttt{convolve\_source\_timefunction.csh} for this purpose,
or alternatively use signal-processing software packages such as SAC \urlwithparentheses{www.llnl.gov/sac}.
Type

\begin{lyxcode}
make~convolve\_source\_timefunction
\end{lyxcode}
to compile the code and then set the parameter \texttt{hdur} in \texttt{convolve\_source\_timefunction.csh}
to the desired half-duration.

@@

\texttt{make generate\_databases}

\texttt{make specfem3D} \\

@@

The files are in a fairly complicated
binary format, but there is a program provided to convert the
output into more user friendly formats:
\begin{description}
\item [\texttt{xcreate\_movie\_shakemap\_AVS\_DX\_GMT}]
From \texttt{create\_movie\_shakemap\_AVS\_DX\_GMT.f90}, it
outputs data in ASCII, OpenDX, or AVS format (also readable in ParaView).
Before compiling the code, make sure you have the file \texttt{surface\_from\_mesher.h}
in the  \texttt{in\_out\_files/OUTPUT\_FILES/} directory. This file will be created by the solver run.
Then type `\texttt{make} \texttt{create\_movie\_shakemap\_AVS\_DX\_GMT}'
and run the executable \texttt{xcreate\_movie\_shakemap\_AVS\_DX\_GMT}
in the \texttt{bin/} subdirectory. It will create visualization files in your format of choice. The code
will prompt the user for input parameters.

@@

To visualize these files, we use an auxilliary program \texttt{combine\_vol\_data.f90}
to combine the data from all slices into one mesh file. To compile it in the root directory, type:
\begin{lyxcode}
{\footnotesize make~combine\_vol\_data~}{\footnotesize \par}
\end{lyxcode}
which will create the executable \texttt{xcombine\_vol\_data} in the directory \texttt{bin/}. To output the
usage of this executable, type `./xcombine\_vol\_data` without arguments.
As an example, you would use
\begin{lyxcode}
{\footnotesize cd~bin/~}{\footnotesize \par}
{\footnotesize ./xcombine\_vol\_data~0~3~velocity\_Z\_it000400~../in\_out\_files/DATABASES\_MPI~../in\_out\_files/OUTPUT\_FILES~0~}{\footnotesize \par}
\end{lyxcode}
to create a low-resolution mesh file. The output mesh file will have the name \texttt{velocity\_Z\_it000400.mesh}.
We next convert the \texttt{.mesh} file into the VTU (Unstructured
grid file) format which can be viewed in ParaView. For this task, you can use and modify the script \texttt{mesh2vtu.pl}
located in directory \texttt{utils/Visualization/Paraview/}, for example:
\begin{lyxcode}
{\footnotesize mesh2vtu.pl~-i~velocity\_Z\_it000400.mesh~-o~velocity\_Z\_it000400.vtu}{\footnotesize \par}
\end{lyxcode}
Notice that this Perl script uses a program \texttt{mesh2vtu} in the
\texttt{utils/Visualization/Paraview/mesh2vtu} directory, which further uses the VTK \urlwithparentheses{http://www.vtk.org/}
run-time library for its execution. Therefore, make sure you have
them properly set in the script according to your system.

@@

\begin{lyxcode}
{\footnotesize make~combine\_vol\_data~}{\footnotesize \par}

@@

\begin{lyxcode}
mesh2vtu.pl~-i~file.mesh~-o~file.vtu
\end{lyxcode}
\item Notice that this Perl script uses a program \texttt{mesh2vtu} in the
\texttt{utils/Visualization/Paraview/mesh2vtu} directory, which further uses the VTK \urlwithparentheses{http://www.vtk.org/}
run-time library for its execution. Therefore, make sure you have
them properly set in the script according to your system.
\end{enumerate}
\item \textbf{Copy over the source and receiver .vtk file}

@@

make~meshfem3D~

make~generate\_databases~

make~specfem3D~

@@

This script converts the synthetic output from the SEM code from ASCII
to SAC format, and performs similar operations as `\texttt{process\_data.pl}'.
An example of the usage of the script:

\begin{lyxcode}
{\footnotesize process\_syn.pl~-m~CMTSOLUTION~-h~-a~STATIONS~-s~1.0~-l~0/4000~-f~-t~40/500~-p~-x~bp~SEM/{*}.BX?.semd}
{\footnotesize \par}
\end{lyxcode}
which will convolve the synthetics with a triangular source-time function
from the \texttt{CMTSOLUTION} file, convert the synthetics into SAC
format, add event and station information into the SAC headers, resample the SAC files with a sampling rate of 1 seconds, cut them between 0 and 4000 seconds, filter them between 40 and
500 seconds with the same filter used for the observed data, pick the first P and S arrivals, and add the suffix `\texttt{bp}'
to the file names.

@@

A sample program \texttt{remap\_database} is provided to map the local
database from a set of machines to another set of machines. This is
especially useful when you want to run mesher and solver, or different
types of solvers separately through a scheduler (refer to Chapter~\ref{cha:Scheduler}).

\begin{lyxcode}
run\_lsf.bash~-{}-gm-no-shmem~-{}-gm-copy-env~remap\_database~old\_machines~150
\end{lyxcode}
where \texttt{old\_machines} is the LSF machine file used in the previous
simulation, and \texttt{150} is the number of processors in total.

>>>>

The SPECFEM3D software package comes in a gzipped tar ball. In the
directory in which you want to install the package, type
\begin{lyxcode}
{\small tar~-zxvf~SPECFEM3D\_V2.0.0.tar.gz}{\small \par}
\end{lyxcode}
The directory \texttt{\small SPECFEM3D/} will then contain
the source code. To configure the software for your system, run the
\texttt{configure} shell script. This script will attempt to guess
the appropriate configuration values for your system. However, at
a minimum, it is recommended that you explicitly specify the appropriate
command names for your Fortran90 compiler and MPI package:

@@

\subsubsection*{Checking the mesh quality}
The quality of the mesh may be inspected more precisely based upon the serial code
in the file  \texttt{check\_mesh\_quality\_}~\\
\texttt{CUBIT\_Abaqus.f90} located in the directory \texttt{src/check\_mesh\_quality\_CUBIT\_Abaqus/}.
Running this code is optional because no information needed by the
solver is generated.

Prior to running and compiling this code, you have to export your mesh in CUBIT to an ABAQUS (.inp) format.
For example, export mesh block IDs belonging to volumes in order to check the quality of the hexahedral elements.
You also have to determine a number of parameters of your mesh, such as the number of nodes and number of elements
and modify the header of the \texttt{check\_mesh\_quality\_CUBIT\_Abaqus.f90} source file. Then,
in the main directory, type
\begin{lyxcode}
{\small make xcheck\_mesh\_quality\_CUBIT\_Abaqus}{\small \par}
\end{lyxcode}
and use
\begin{lyxcode}
{\small ./bin/xcheck\_mesh\_quality\_CUBIT\_Abaqus}{\small \par}
\end{lyxcode}
to generate an AVS output file (\texttt{\small AVS\_meshquality.inp}
in AVS UCD format) or OpenDX output file (\texttt{\small DX\_meshquality.}~\\
\texttt{\small dx}) that can be used to investigate mesh quality,
e.g, skewness of elements and a Gnuplot histogram (\texttt{\small mesh\_quality\_}~\\
\texttt{\small histogram.txt}) that can be plotted with gnuplot (type
`\texttt{\small gnuplot plot\_mesh\_quality\_histogram.gnu}'). The
histogram is also printed to the screen.
Analyze that skewness histogram of mesh elements to make sure no element has a skewness above approximately 0.75, otherwise the mesh is of poor quality (and if even a single element has a skewness value above 0.80, then you must definitely improve the mesh).
If you want to start designing
your own meshes, this tool is useful for viewing your creations. You
are striving for meshes with elements with `cube-like' dimensions,
e.g., the mesh should contain no very elongated or skewed elements.

@@

When you are ready to compile,
in the main directory type
\begin{lyxcode}
{\small make xdecompose\_mesh\_SCOTCH}{\small \par}
\end{lyxcode}
If all paths and flags have been set correctly, the executable \texttt{bin/xdecompose\_mesh\_SCOTCH} should be produced.

@@

In the main directory, type 
\begin{lyxcode}
{\small make xmeshfem3D}{\small \par}
\end{lyxcode}
If all paths and flags
have been set correctly, the mesher should now compile and produce
the executable \texttt{bin/xmeshfem3D}. Please note that
\texttt{xmeshfem3D} must be called directly from the \texttt{bin/} directory,
as most of the binaries of the package.

@@

After using \texttt{xmeshfem3D} or \texttt{xdecompose\_mesh\_SCOTCH}, the next step in the workflow is to compile 
\texttt{xgenerate\_}~\\
\texttt{databases}. This program is going to create all the missing information needed by the SEM solver.

@@

In the main directory, type 
\begin{lyxcode}
{\small make xgenerate\_databases}{\small \par}
\end{lyxcode}
Input for the program is provided through
the main parameter file \texttt{Par\_file}, which resides in the subdirectory \texttt{in\_data\_files/}. Please note that
\texttt{xgenerate\_databases} must be called directly from the \texttt{bin/} directory, as most of the binaries of the package.

@@

Now that you have successfully generated the databases, you are ready to compile
the solver. In the main directory, type
\begin{lyxcode}
{\small make xspecfem3D}{\small \par}
\end{lyxcode}
Please note that \texttt{xspecfem3D} must be called directly from
the \texttt{bin/} directory, as most of the binaries of the package.

@@

\begin{itemize}
\item Set the \texttt{time shift} parameter equal to $0.0$ (the solver
will not run otherwise.) The time shift parameter would simply apply
an overall time shift to the synthetics, something that can be done
in the post-processing (see Section \ref{sec:Process-data-and-syn}).
\item For point-source simulations (see finite sources, page \pageref{To-simulate-a})
we recommend setting the source half-duration parameter \texttt{half
duration} equal to zero, which corresponds to simulating a step source-time
function, i.e., a moment-rate function that is a delta function. If
\texttt{half duration} is not set to zero, the code will use a Gaussian
(i.e., a signal with a shape similar to a `smoothed triangle', as
explained in \citet{KoTr02a} and shown in Fig~\ref{fig:gauss.vs.triangle})
source-time function with half-width \texttt{half duration}. We prefer
to run the solver with \texttt{half duration} set to zero and convolve
the resulting synthetic seismograms in post-processing after the run,
because this way it is easy to use a variety of source-time functions
(see Section \ref{sec:Process-data-and-syn}). \citet{KoTr02a} determined
that the noise generated in the simulation by using a step source
time function may be safely filtered out afterward based upon a convolution
with the desired source time function and/or low-pass filtering. Use
the serial code \texttt{convolve\_source\_timefunction.f90} and the
script \texttt{convolve\_source\_timefunction.csh} for this purpose,
or alternatively use signal-processing software packages such as SAC \urlwithparentheses{www.llnl.gov/sac}.
Type
\begin{lyxcode}
{\small make xconvolve\_source\_timefunction}{\small \par}
\end{lyxcode}
to compile the code and then set the parameter \texttt{hdur} in \texttt{convolve\_source\_timefunction.csh}
to the desired half-duration.

@@

\texttt{make xgenerate\_databases}

\texttt{make xspecfem3D} \\

@@

The files are in a fairly complicated
binary format, but there is a program provided to convert the
output into more user friendly formats:
\begin{description}
\item [\texttt{xcreate\_movie\_shakemap\_AVS\_DX\_GMT}]
From \texttt{create\_movie\_shakemap\_AVS\_DX\_GMT.f90}, it
outputs data in ASCII, OpenDX, or AVS format (also readable in ParaView).
Before compiling the code, make sure you have the file \texttt{surface\_from\_mesher.h}
in the  \texttt{in\_out\_files/OUTPUT\_FILES/} directory. This file will be created by the solver run.
Then type 
\begin{lyxcode}
{\footnotesize make xcreate\_movie\_shakemap\_AVS\_DX\_GMT}{\footnotesize \par}
\end{lyxcode}
and run the executable \texttt{xcreate\_movie\_shakemap\_AVS\_DX\_GMT}
in the \texttt{bin/} subdirectory. It will create visualization files in your format of choice. The code
will prompt the user for input parameters.

@@

To visualize these files, we use an auxilliary program \texttt{combine\_vol\_data.f90}
to combine the data from all slices into one mesh file. To compile it in the root directory, type:
\begin{lyxcode}
{\footnotesize make~xcombine\_vol\_data~}{\footnotesize \par}
\end{lyxcode}
which will create the executable \texttt{xcombine\_vol\_data} in the directory \texttt{bin/}. To output the
usage of this executable, type `./bin/xcombine\_vol\_data` without arguments.
As an example, to run the executable you would use
\begin{lyxcode}
{\footnotesize cd~bin/~}{\footnotesize \par}
{\footnotesize ./xcombine\_vol\_data~0~3~velocity\_Z\_it000400~../in\_out\_files/DATABASES\_MPI~../in\_out\_files/OUTPUT\_FILES~0~}{\footnotesize \par}
\end{lyxcode}
to create a low-resolution mesh file. The output mesh file will have the name \texttt{velocity\_Z\_it000400.mesh}.
We next convert the \texttt{.mesh} file into the VTU (Unstructured
grid file) format which can be viewed in ParaView. For this task, you can use and modify the script \texttt{mesh2vtu.pl}
located in directory \texttt{utils/Visualization/Paraview/}, for example:
\begin{lyxcode}
{\footnotesize mesh2vtu.pl~-i~velocity\_Z\_it000400.mesh~-o~velocity\_Z\_it000400.vtu}{\footnotesize \par}
\end{lyxcode}
Notice that this Perl script uses a program \texttt{mesh2vtu} in the
\texttt{utils/Visualization/Paraview/mesh2vtu} directory, which further uses the VTK \urlwithparentheses{http://www.vtk.org/}
run-time library for its execution. Therefore, make sure you have
them properly set in the script according to your system.

@@

\begin{lyxcode}
{\footnotesize make~xcombine\_vol\_data~}{\footnotesize \par}

@@

\begin{lyxcode}
{\footnotesize mesh2vtu.pl~-i~file.mesh~-o~file.vtu~}{\footnotesize \par}
\end{lyxcode}
\item Notice that this Perl script uses a program \texttt{mesh2vtu} in the
\texttt{utils/Visualization/Paraview/mesh2vtu} directory, which further uses the VTK \urlwithparentheses{http://www.vtk.org/}
run-time library for its execution. Therefore, make sure you have
them properly set in the script according to your system.
\end{enumerate}
\item \textbf{Copy over the source and receiver .vtk file}

@@

make~xmeshfem3D~

make~xgenerate\_databases~

make~xspecfem3D~

@@

This script converts the synthetic output from the SEM code from ASCII
to SAC format, and performs similar operations as `\texttt{process\_data.pl}'.
An example of the usage of the script:
\begin{lyxcode}
{\footnotesize process\_syn.pl~-m~CMTSOLUTION~-h~-a~STATIONS~-s~1.0~-l~0/4000~-f~-t~40/500~-p~-x~bp~SEM/{*}.BX?.semd}
{\footnotesize \par}
\end{lyxcode}
which will convolve the synthetics with a triangular source-time function
from the \texttt{CMTSOLUTION} file, convert the synthetics into SAC
format, add event and station information into the SAC headers, resample the SAC files with a sampling rate of 1 seconds, cut them between 0 and 4000 seconds, filter them between 40 and
500 seconds with the same filter used for the observed data, pick the first P and S arrivals, and add the suffix `\texttt{bp}'
to the file names.

@@

A sample program \texttt{remap\_database} is provided to map the local
database from a set of machines to another set of machines. This is
especially useful when you want to run mesher and solver, or different
types of solvers separately through a scheduler (refer to Chapter~\ref{cha:Scheduler}).
\begin{lyxcode}
run\_lsf.bash~-{}-gm-no-shmem~-{}-gm-copy-env~remap\_database~old\_machines~150
\end{lyxcode}
where \texttt{old\_machines} is the LSF machine file used in the previous
simulation, and \texttt{150} is the number of processors in total.
